---
title: "Multiscale Aggregated EGPD Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiscale Aggregated EGPD Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

The `egpd` package includes compound Poisson EGPD models for a **single**
aggregation scale through `dcpegpd()`, `dcpdegpd()`, and
`distfit(..., family = "cpegpd"/"cpdegpd")`. The
`fit_aggregate_egpd()` function extends that infrastructure to a
**multiscale** setting, following the aggregated EGPD framework described by
Ailliot, Gaetan, and Naveau (2026).

The current implementation fits the positive part of the model across several
durations at once:

$$
A_d \sim \mathrm{EGPD}(\sigma_d, \kappa, \xi, \lambda_d),
$$

with shared shape parameters $\kappa$ and $\xi$, and duration-specific
parameters modeled as polynomials in $\log d$:

$$
\log \sigma_d = s_0 + s_1 \log d + \cdots + s_p (\log d)^p,
\qquad
\log \lambda_d = l_0 + l_1 \log d + \cdots + l_q (\log d)^q.
$$

Here, $\lambda_d$ controls the zero mass through
$P(A_d = 0) = \exp(-\lambda_d)$, while the positive aggregates are modeled
through the compound Poisson EGPD law. In the current package implementation,
`predict(..., type = "quantile")` returns **positive-part quantiles**, that
is, quantiles conditional on the aggregate being strictly positive.

## 1. A synthetic multiscale example

We start from a set of aggregated samples already grouped by duration. This is
the cleanest way to see the fitted multiscale model because the data are
generated directly from the family used by `fit_aggregate_egpd()`.

```{r multiscale-sim}
library(egpd)
set.seed(2)

durations <- c(1L, 2L, 4L, 8L)
sigma_coef_true <- c(log(0.6), 0.12)
lambda_coef_true <- c(log(0.8), 0.25)
kappa_true <- 0.35
xi_true <- 0.2

sigma_true <- exp(sigma_coef_true[1] + sigma_coef_true[2] * log(durations))
lambda_true <- exp(lambda_coef_true[1] + lambda_coef_true[2] * log(durations))

agg_samples <- setNames(lapply(seq_along(durations), function(i) {
  rcpegpd(
    300,
    sigma = sigma_true[i],
    xi = xi_true,
    kappa = kappa_true,
    lambda = lambda_true[i],
    type = 1
  )
}), durations)

data.frame(
  duration = durations,
  n = vapply(agg_samples, length, integer(1)),
  prop_zero = round(vapply(agg_samples, function(x) mean(x == 0), numeric(1)), 3),
  mean = round(vapply(agg_samples, mean, numeric(1)), 2)
)
```

The aggregate samples show the two main ingredients of the model:

1. zeros become less frequent as $\lambda_d$ increases with duration;
2. positive aggregates typically get larger as $\sigma_d$ increases.

## 2. Fitting the multiscale model

The fit shares `kappa` and `xi` across durations and estimates the polynomial
coefficients for `log(sigma_d)` and `log(lambda_d)`. In this example we use
linear trends in `log(d)`, so `p = 1` and `q = 1`.

```{r multiscale-fit}
fit_agg <- fit_aggregate_egpd(
  agg_samples,
  durations = durations,
  family = "cpegpd",
  p = 1,
  q = 1,
  start = list(
    sigma_coef = sigma_coef_true,
    lambda_coef = lambda_coef_true,
    kappa = kappa_true,
    xi = xi_true
  ),
  optim.method = "BFGS",
  control = list(maxit = 200)
)

fit_agg
```

```{r multiscale-summary}
summary(fit_agg)
```

The fitted object stores both the regression-style coefficients and the
duration-specific parameter values implied by those coefficients.

```{r multiscale-parameters}
fit_pars <- predict(fit_agg, type = "parameter")

comparison <- data.frame(
  duration = durations,
  sigma_true = sigma_true,
  sigma_hat = fit_pars$sigma,
  lambda_true = lambda_true,
  lambda_hat = fit_pars$lambda,
  kappa_true = kappa_true,
  kappa_hat = fit_pars$kappa,
  xi_true = xi_true,
  xi_hat = fit_pars$xi
)

round(comparison, 3)
```

## 3. Extrapolating across durations

Because the scale and Poisson mean are modeled as smooth functions of
$\log d$, we can predict parameters and positive-part quantiles at durations
that were not used directly in the fit.

```{r multiscale-predict}
new_durations <- c(1L, 2L, 4L, 8L, 16L)

round(
  predict(fit_agg, durations = new_durations, type = "parameter"),
  3
)
```

```{r multiscale-quantiles}
round(
  predict(
    fit_agg,
    durations = new_durations,
    type = "quantile",
    prob = c(0.5, 0.9, 0.99)
  ),
  2
)
```

The monotone parameter curves are visible in the default plot method:

```{r multiscale-plot, fig.height = 4}
plot(fit_agg)
```

## 4. Starting from a finest-scale series

If you have the original non-negative series rather than precomputed
aggregates, `fit_aggregate_egpd()` can construct the duration-specific sums
internally.

```{r series-fit, eval = FALSE}
set.seed(4)
x <- rcpegpd(300, sigma = 0.8, xi = 0.2, kappa = 0.35, lambda = 0.7, type = 1)

fit_from_series <- fit_aggregate_egpd(
  x,
  durations = c(1L, 2L, 4L, 8L),
  family = "cpegpd",
  p = 1,
  q = 1,
  overlap = TRUE
)

predict(fit_from_series, type = "parameter")
```

With `overlap = TRUE`, the function uses rolling sums. Set `overlap = FALSE`
to use non-overlapping block sums instead.

## 5. Discrete aggregates

For integer-valued aggregated counts, switch to `family = "cpdegpd"`. The
interface is the same, but the aggregated samples must be integer-valued.

```{r discrete-fit, eval = FALSE}
set.seed(3)
count_samples <- setNames(lapply(durations, function(d) {
  rcpdegpd(
    250,
    sigma = exp(log(1.2) + 0.08 * log(d)),
    xi = 0.18,
    kappa = 0.4,
    lambda = exp(log(0.9) + 0.22 * log(d)),
    type = 1
  )
}), durations)

fit_counts <- fit_aggregate_egpd(
  count_samples,
  durations = durations,
  family = "cpdegpd",
  p = 1,
  q = 1
)

predict(fit_counts, type = "quantile", prob = c(0.5, 0.9))
```

## 6. Current scope

`fit_aggregate_egpd()` is intended as a first package implementation of the
multiscale framework. In particular:

- it uses the package's existing compound Poisson EGPD and compound Poisson
  discrete EGPD building blocks;
- it fits a composite likelihood over **positive** aggregates, while
  `lambda_d` captures the zero mass;
- it enforces monotone fitted curves for `sigma_d` and `lambda_d` during
  optimization to preserve non-crossing behavior across durations.

For single-scale aggregate modeling, or when you need the full single-duration
density/CDF/quantile utilities directly, continue to use `distfit()` together
with `dcpegpd()`, `pcpegpd()`, `qcpegpd()`, `dcpdegpd()`, `pcpdegpd()`, and
`qcpdegpd()`.
