---
title: "Bell Distribution Models for Overdispersed Counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bell Distribution Models for Overdispersed Counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

The Bell distribution of Castellares, Ferrari and Lemonte (2018) is a
*one-parameter* count family,

$$\Pr(Y = y) = \frac{\theta^{y}\,e^{1 - e^{\theta}}\,B_y}{y!},
  \qquad y = 0, 1, 2, \ldots,$$

where $B_y$ are the Bell numbers. It has mean $\theta e^{\theta}$ and variance
$\theta(1+\theta)e^{\theta}$, so its index of dispersion is $1 + \theta > 1$:
the Bell is always overdispersed relative to the Poisson, making it a natural
one-parameter alternative when the Poisson's equidispersion is too restrictive.
Because there is a single parameter, that overdispersion is tied to the mean
(the dispersion index is $1 + W_0(\mu)$ at mean $\mu$); data with very strong
overdispersion need a heavier model such as the PIG or GPIG.

`egpd` provides two parameterisations: the **mean** parameterisation
(`family = "bell"`) with `mu` (log link) the mean, and the **native**
parameterisation (`family = "bellnat"`) with the paper's `theta` (log link).
The zero-inflated Bell (`"zibell"`) adds a point mass at zero.

## Data

```{r data}
library(egpd)
data(nsw_offenses)
y <- nsw_offenses$offenses
cat("n =", length(y), "  mean =", round(mean(y), 2),
    "  var =", round(var(y), 1), "\n")
cat("proportion of zeros:", round(mean(y == 0), 3),
    "  dispersion (var/mean):", round(var(y) / mean(y), 1), "\n")
```

```{r eda}
plot(table(y), main = "Gaming and betting offenses (NSW)",
     xlab = "Number of offenses", ylab = "Frequency")
```

## Fitting Bell

```{r fits}
df <- data.frame(y = y)

fit_bell  <- egpd(list(lmu = y ~ 1), data = df, family = "bell")
fit_zib   <- egpd(list(lmu = y ~ 1, logitpi = ~ 1), data = df, family = "zibell")

## drop = FALSE keeps a one-row data frame even for the single-parameter Bell
p_bell <- predict(fit_bell, type = "response")[1, , drop = FALSE]
p_zib  <- predict(fit_zib,  type = "response")[1, , drop = FALSE]
p_bell
```

The Bell parameter and the implied dispersion index follow from the fitted
mean:

```{r theta}
theta_hat <- as.numeric(egpd:::bell_W0_cpp(p_bell$mu))   # theta = W0(mu)
cat("fitted mu =", round(p_bell$mu, 2),
    "  theta =", round(theta_hat, 3),
    "  Bell dispersion index (1 + theta) =", round(1 + theta_hat, 2), "\n")
cat("empirical dispersion (var/mean)     =", round(var(y) / mean(y), 2), "\n")
```

The Bell captures the overdispersion that the Poisson cannot, but its
dispersion index is bounded by the mean, so it explains only part of the excess
variance here.

## Model comparison

```{r compare}
fit_pois <- glm(y ~ 1, family = poisson)
fit_pig  <- egpd(list(lmu = y ~ 1, lsigma = ~ 1), data = df, family = "pig")

comparison <- data.frame(
  model  = c("Poisson", "Bell", "ZIBell", "PIG"),
  npar   = c(1, 1, 2, 2),
  logLik = c(as.numeric(logLik(fit_pois)), as.numeric(logLik(fit_bell)),
             as.numeric(logLik(fit_zib)),  as.numeric(logLik(fit_pig))),
  AIC    = c(AIC(fit_pois), AIC(fit_bell), AIC(fit_zib), AIC(fit_pig))
)
comparison
```

The single-parameter Bell already improves substantially on the Poisson; the
two-parameter PIG, which decouples dispersion from the mean, fits the strong
overdispersion better still.

## Fitted versus empirical PMF

```{r pmf, fig.width = 7, fig.height = 5}
xv <- 0:max(y)
emp <- tabulate(y + 1, nbins = max(y) + 1) / length(y)

pmf_bell <- dBELL(xv, mu = p_bell$mu)
pmf_pois <- dpois(xv, mean(y))

plot(xv, emp, type = "h", lwd = 2, col = "grey70",
     main = "Empirical vs fitted PMF", xlab = "Offenses", ylab = "Probability")
lines(xv + 0.2, pmf_bell, type = "h", lwd = 2, col = "steelblue")
lines(xv + 0.4, pmf_pois, type = "h", lwd = 2, col = "darkorange")
legend("topright", c("Empirical", "Bell", "Poisson"), lwd = 2,
       col = c("grey70", "steelblue", "darkorange"))
```

The Bell spreads mass into the tail where the Poisson cannot.

## Randomized quantile residuals

```{r rqr, fig.width = 7, fig.height = 4}
rqr <- function(y, pfun) {
  Fup <- pfun(y); Flo <- pfun(y - 1); Flo[y == 0] <- 0
  qnorm(runif(length(y), Flo, Fup))
}

set.seed(1)
r_bell <- rqr(y, function(q) pBELL(q, mu = p_bell$mu))
r_zib  <- rqr(y, function(q) pZIBELL(q, mu = p_zib$mu, sigma = p_zib$pi))

par(mfrow = c(1, 2))
qqnorm(r_bell, main = "Bell", pch = 20, col = "grey50"); qqline(r_bell, col = "red")
qqnorm(r_zib,  main = "ZIBell", pch = 20, col = "grey50"); qqline(r_zib, col = "red")
par(mfrow = c(1, 1))
```

## Other fitting interfaces

The univariate maximum-likelihood fitter `fitegpd()` uses the native `theta`
parameterisation, and reaches the closed-form estimate
$\hat{\theta} = W_0(\bar{y})$:

```{r other}
fit_uni <- fitegpd(y, family = "bell")
cat("fitegpd theta =", round(unname(fit_uni$estimate["theta"]), 4),
    "  W0(ybar) =", round(as.numeric(egpd:::bell_W0_cpp(mean(y))), 4), "\n")
```

`BELL()` / `ZIBELL()` (`gamlss`) and `bell_bamlss()` / `zibell_bamlss()`
(`bamlss`) expose the Bell for distributional regression, and `dbell`, `pbell`,
`qbell`, `rbell` (plus the `zibell` variants) give the density, distribution,
quantile and simulation functions. For heavier tails than the Bell can reach,
see `vignette("pig-counts")` and `vignette("gpig-counts")`.
