---
title: "Composite Pareto Distributions and GAMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composite Pareto Distributions and GAMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

The `egpd` package includes continuous composite Pareto distribution helpers
and a `family = "comppareto"` option for `egpd()`. This follows the composite
construction used in the CRAN package
[`CompPareto`](https://CRAN.R-project.org/package=CompPareto), where a
continuous body distribution is joined to a Pareto tail at a positive splice
point `theta`.

The current `egpd` implementation covers two related tasks:

1. unconditional `d/p/q/rcomppareto()` utilities for continuous composite
   Pareto distributions;
2. additive CompPareto regression through the existing `egpd()` GAM framework.

## 1. The composite Pareto construction

Let `G1` and `g1` denote the body CDF and density. In the current package
version, the body can be one of:

- `"lnorm"`
- `"gamma"`
- `"weibull"`
- `"exp"`

Above the splice point `theta`, the model uses a Pareto tail with shape
parameter `alpha`. The weight is chosen so that the density is continuous at
the splice point. In practice, this gives a model that behaves like the chosen
body below `theta` and like a heavy-tailed Pareto model above `theta`.

## 2. Standalone distribution utilities

We start with a lognormal body. The standalone helpers use the same
parameterization as the body distribution plus the Pareto tail parameters
`alpha` and `theta`.

```{r standalone-setup}
library(egpd)

args <- list(
  spec = "lnorm",
  meanlog = 0.4,
  sdlog = 0.7,
  alpha = 1.7,
  theta = 3.2
)

probs <- c(0.5, 0.9, 0.99)
qvals <- do.call(qcomppareto, c(list(p = probs), args))

data.frame(
  prob = probs,
  quantile = round(qvals, 3),
  cdf_back = round(do.call(pcomppareto, c(list(q = qvals), args)), 6)
)
```

The returned probabilities agree with the inputs, showing the usual
`p(q(p)) = p` roundtrip.

```{r standalone-curves}
grid <- seq(0.01, 20, length.out = 400)
cdf_vals <- do.call(pcomppareto, c(list(q = grid), args))
dens_vals <- do.call(dcomppareto, c(list(x = grid), args))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(grid, cdf_vals, type = "l", lwd = 2,
     xlab = "y", ylab = "CDF", main = "Composite Pareto CDF")
abline(v = args$theta, lty = 2, col = "grey50")

plot(grid, dens_vals, type = "l", lwd = 2,
     xlab = "y", ylab = "density", main = "Composite Pareto density")
abline(v = args$theta, lty = 2, col = "grey50")
par(op)
```

The dashed line marks the splice point `theta`.

We can also inspect the density numerically just below and just above the
splice point.

```{r splice-continuity}
eps <- 1e-6

data.frame(
  x = c(args$theta - eps, args$theta, args$theta + eps),
  density = do.call(
    dcomppareto,
    c(list(x = c(args$theta - eps, args$theta, args$theta + eps)), args)
  )
)
```

The values are nearly identical, reflecting the continuity built into the
construction.

## 3. Supported body specifications

The unconditional helpers and the GAM family share the same `spec` choices.
The main difference is the naming convention for the regression formulas.

```{r supported-specs}
data.frame(
  spec = c("lnorm", "gamma", "weibull", "exp"),
  standalone_parameters = c(
    "meanlog, sdlog, alpha, theta",
    "shape, scale, alpha, theta",
    "shape, scale, alpha, theta",
    "rate, alpha, theta"
  ),
  gam_formula_names = c(
    "meanlog, logsdlog, logalpha, logtheta",
    "logshape, logscale, logalpha, logtheta",
    "logshape, logscale, logalpha, logtheta",
    "lograte, logalpha, logtheta"
  )
)
```

The exponential-body case is especially convenient for regression examples
because it only needs three parameter formulas.

## 4. Fitting a CompPareto GAM

Here we simulate data from an exponential-body composite Pareto model with a
smooth predictor effect on the body rate. For this illustration, we use a
somewhat larger sample and a higher splice point so the nonlinear body signal
is visible in the fitted GAM.

```{r gam-sim}
set.seed(22)

n <- 400
x <- sort(stats::runif(n, -1, 1))
rate <- exp(-0.2 + 1.0 * sin(pi * x))
y <- vapply(
  seq_len(n),
  function(i) rcomppareto(1, spec = "exp", rate = rate[i], alpha = 1.3, theta = 3.5),
  numeric(1)
)

dat <- data.frame(y = y, x = x)
```

We then fit the composite Pareto GAM by selecting `family = "comppareto"` and
passing the body specification through `comppareto.args`.

```{r gam-fit}
fit <- suppressMessages(
  egpd(
    list(lograte = y ~ s(x, k = 7), logalpha = ~1, logtheta = ~1),
    data = dat,
    family = "comppareto",
    comppareto.args = list(spec = "exp")
  )
)

fit$convergence
```

The fitted response-scale parameters can be predicted in the usual way.

```{r gam-response}
pred_grid <- data.frame(x = seq(-1, 1, length.out = 100))
pars_hat <- predict(fit, newdata = pred_grid, type = "response")
rate_true <- exp(-0.2 + 1.0 * sin(pi * pred_grid$x))

head(round(pars_hat, 3))
```

```{r gam-plot, fig.height = 4}
plot(pred_grid$x, pars_hat$rate, type = "l", lwd = 2,
     xlab = "x", ylab = "rate", main = "Exponential-body rate: true vs fitted")
lines(pred_grid$x, rate_true, lwd = 2, lty = 2, col = "steelblue")
points(x, rate, pch = 16, cex = 0.4, col = grDevices::adjustcolor("black", alpha.f = 0.25))
legend("topright",
       legend = c("fitted", "true", "simulated truth"),
       lty = c(1, 2, NA), lwd = c(2, 2, NA), pch = c(NA, NA, 16),
       col = c("black", "steelblue", grDevices::adjustcolor("black", alpha.f = 0.25)),
       bty = "n")
```

In this setup, the fitted smooth tracks the nonlinear truth well rather than
shrinking toward a nearly linear trend.

Quantile prediction also works through the regular `predict.egpd()` interface.

```{r gam-quantiles}
round(
  predict(
    fit,
    newdata = data.frame(x = c(-0.75, 0, 0.75)),
    type = "quantile",
    prob = c(0.5, 0.9, 0.99)
  ),
  3
)
```

Randomized quantile residuals are available as well.

```{r gam-rqresid}
summary(rqresid(fit, seed = 1))
```

## 5. Intercept-only four-parameter fits

The same interface also supports four-parameter bodies such as the lognormal.
In that case, the formula list simply includes an extra body parameter:

```{r lognormal-formula, eval = FALSE}
fit_lnorm <- egpd(
  list(meanlog = y ~ 1, logsdlog = ~1, logalpha = ~1, logtheta = ~1),
  data = data.frame(y = y),
  family = "comppareto",
  comppareto.args = list(spec = "lnorm")
)
```

This is often the simplest way to start when you want to compare different body
specifications before adding predictor effects.

## 6. Current scope

The present CompPareto implementation in `egpd` is intentionally focused:

1. only the continuous body specifications `"lnorm"`, `"gamma"`, `"weibull"`,
   and `"exp"` are currently exposed;
2. the GAM family uses numerical derivatives on the linear predictor scale, so
   it is usually slower than the native EGPD likelihoods;
3. randomized quantile residuals and quantile prediction are supported for the
   new family;
4. discrete composite Pareto helpers from the upstream `CompPareto` package are
   not yet part of the `egpd()` GAM layer.

That makes the current implementation a good fit for continuous bulk-plus-tail
modeling, while leaving room for broader composite-distribution extensions
later.
