---
title: "Generalised Poisson-Inverse Gaussian Models for Heavy-Tailed Counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalised Poisson-Inverse Gaussian Models for Heavy-Tailed Counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

The generalised Poisson-inverse Gaussian (GPIG) distribution of Zhu and Joe
(2009) is a three-parameter count family with probability generating function
$G(s) = \exp\{b[(1-c)^a - (1-cs)^a]\}$, for $0 < a \le 1$, $b > 0$ and
$0 \le c \le 1$. It nests several familiar models:

* $a = 1$ gives the **Poisson** (mean $bc$),
* $a = 1/2$ gives the **Poisson-inverse Gaussian** (see
  `vignette("pig-counts")`),
* $c = 1$ gives the **discrete-stable** distribution.

The tail exponent $a$ controls how heavy the right tail is: smaller $a$ means a
heavier tail than the PIG can produce. `egpd` offers two parameterisations. The
**mean** parameterisation (used by `family = "gpig"`) models $\mathbb{E}(Y) =
\mu$ directly with `mu` (log link), `a` (logit link) and `c` (logit link); the
**native** $(a, b, c)$ parameterisation (`family = "gpignat"`) is faithful to
the paper. This vignette fits GPIG to a strongly heavy-tailed count and
compares it with the Poisson and PIG.

## Data

```{r data}
library(egpd)
data(ny_complaints)
y <- ny_complaints$upheld
cat("n =", length(y), "  mean =", round(mean(y), 2),
    "  var =", round(var(y), 1), "  max =", max(y), "\n")
cat("proportion of zeros:", round(mean(y == 0), 3),
    "  dispersion (var/mean):", round(var(y) / mean(y), 1), "\n")
```

```{r eda}
plot(table(y[y <= 40]), main = "Upheld insurance complaints (NY)",
     xlab = "Number of upheld complaints", ylab = "Frequency")
```

The dispersion index exceeds 40 and the counts reach the hundreds: a very heavy
tail that the negative binomial or PIG struggle with.

## Fitting GPIG and ZIGPIG

```{r fits}
df <- data.frame(y = y)

fit_gpig  <- egpd(list(lmu = y ~ 1, logita = ~ 1, logitc = ~ 1),
                  data = df, family = "gpig")
fit_zig   <- egpd(list(lmu = y ~ 1, logita = ~ 1, logitc = ~ 1, logitpi = ~ 1),
                  data = df, family = "zigpig")

p_gpig <- predict(fit_gpig, type = "response")[1, ]
p_zig  <- predict(fit_zig,  type = "response")[1, ]
p_gpig
```

The GPIG places the heaviness of the tail in two parameters: the tail exponent
`a` and the down-weighting `c`. Here `c` is pushed close to $1$ (the
discrete-stable regime, which is where the GPIG's heaviest tails live) while
`a` sits near the PIG value of $1/2$.

```{r ahat}
cat("fitted a (tail exponent):", round(p_gpig$a, 3),
    "  fitted c (down-weighting):", round(p_gpig$c, 3),
    "\n  [PIG corresponds to a = 0.5; c -> 1 gives the discrete-stable tail]\n")
```

## Model comparison

```{r compare}
fit_pois <- glm(y ~ 1, family = poisson)
fit_pig  <- egpd(list(lmu = y ~ 1, lsigma = ~ 1), data = df, family = "pig")

comparison <- data.frame(
  model  = c("Poisson", "PIG", "GPIG", "ZIGPIG"),
  npar   = c(1, 2, 3, 4),
  logLik = c(as.numeric(logLik(fit_pois)), as.numeric(logLik(fit_pig)),
             as.numeric(logLik(fit_gpig)), as.numeric(logLik(fit_zig))),
  AIC    = c(AIC(fit_pois), AIC(fit_pig), AIC(fit_gpig), AIC(fit_zig))
)
comparison
```

Both mixed-Poisson families dwarf the Poisson. The extra tail flexibility of
the GPIG gives a further improvement over the PIG, and zero-inflation adds a
little more again.

## Fitted versus empirical PMF

```{r pmf, fig.width = 7, fig.height = 5}
xv <- 0:40
emp <- tabulate(y + 1, nbins = max(y) + 1)[seq_along(xv)] / length(y)

pmf_gpig <- dGPIG(xv, mu = p_gpig$mu, sigma = p_gpig$a, nu = p_gpig$c)
pmf_zig  <- dZIGPIG(xv, mu = p_zig$mu, sigma = p_zig$a, nu = p_zig$c,
                    tau = p_zig$pi)

plot(xv, emp, type = "h", lwd = 3, col = "grey70",
     main = "Empirical vs fitted PMF", xlab = "Upheld complaints",
     ylab = "Probability")
lines(xv + 0.15, pmf_gpig, type = "h", lwd = 2, col = "steelblue")
lines(xv + 0.30, pmf_zig,  type = "h", lwd = 2, col = "firebrick")
legend("topright", c("Empirical", "GPIG", "ZIGPIG"), lwd = 2,
       col = c("grey70", "steelblue", "firebrick"))
```

## Randomized quantile residuals

```{r rqr, fig.width = 7, fig.height = 4}
r_gpig <- rqresid(fit_gpig, seed = 1)
r_zig  <- rqresid(fit_zig,  seed = 1)

par(mfrow = c(1, 2))
qqnorm(r_gpig, main = "GPIG", pch = 20, col = "grey50"); qqline(r_gpig, col = "red")
qqnorm(r_zig,  main = "ZIGPIG", pch = 20, col = "grey50"); qqline(r_zig, col = "red")
par(mfrow = c(1, 1))
```

## Native parameterisation and other interfaces

`family = "gpignat"` fits the paper's native $(a, b, c)$ parameterisation; it
describes the same fitted distribution as `"gpig"` (identical log-likelihood)
but reports $b$ instead of $\mu$. The univariate fitter `distfit()` uses the
native parameterisation directly:

```{r other}
fit_uni <- distfit(y, family = "gpig")
round(fit_uni$estimate, 3)
```

`GPIG()` / `ZIGPIG()` (`gamlss`) and `gpig_bamlss()` / `zigpig_bamlss()`
(`bamlss`) expose the family for distributional regression, and `dgpig`,
`pgpig`, `qgpig`, `rgpig` (plus the `zigpig` variants) give the density,
distribution, quantile and simulation functions.
