---
title: "Poisson-Inverse Gaussian Models for Overdispersed Counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson-Inverse Gaussian Models for Overdispersed Counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

The Poisson-inverse Gaussian (PIG) distribution is a mixed-Poisson model: a
Poisson mean that is itself inverse-Gaussian distributed. The result is a
count distribution that accommodates far more overdispersion than the Poisson,
with a heavier right tail than the negative binomial. In `egpd` the mean
parameterisation is used,

$$\mathbb{E}(Y) = \mu, \qquad \mathrm{Var}(Y) = \mu + \sigma\,\mu^2,$$

so `mu` (log link) is the mean and `sigma` (log link) controls overdispersion.
The zero-inflated variant (ZIPIG) adds a point mass at zero of probability
`pi` (logit link) for data with more zeros than the PIG alone predicts.

This vignette fits PIG and ZIPIG to doctor-visit counts and compares them with
a Poisson baseline.

## Data

```{r data}
library(egpd)
data(docvisits)
y <- docvisits$visits
cat("n =", length(y), "  mean =", round(mean(y), 2),
    "  var =", round(var(y), 1), "\n")
cat("proportion of zeros:", round(mean(y == 0), 3),
    "  dispersion (var/mean):", round(var(y) / mean(y), 1), "\n")
```

```{r eda}
plot(table(y[y <= 30]), main = "Doctor visits",
     xlab = "Number of visits", ylab = "Frequency")
```

The variance is roughly nine times the mean and 41% of the observations are
zero: the Poisson is clearly inadequate.

## Fitting with `egpd()`

The native GAM fitter takes a list of parameter formulae. For intercept-only
models every parameter gets `~ 1`.

```{r fits}
df <- data.frame(y = y)

fit_pig  <- egpd(list(lmu = y ~ 1, lsigma = ~ 1),
                 data = df, family = "pig")
fit_zip  <- egpd(list(lmu = y ~ 1, lsigma = ~ 1, logitpi = ~ 1),
                 data = df, family = "zipig")

summary(fit_pig)
```

The fitted parameters are recovered on the response scale with `predict()`.

```{r pars}
p_pig <- predict(fit_pig, type = "response")[1, ]
p_zip <- predict(fit_zip, type = "response")[1, ]
p_pig
p_zip
```

## Comparison with the Poisson

```{r poisson}
fit_pois <- glm(y ~ 1, family = poisson)

comparison <- data.frame(
  model  = c("Poisson", "PIG", "ZIPIG"),
  npar   = c(1, 2, 3),
  logLik = c(as.numeric(logLik(fit_pois)),
             as.numeric(logLik(fit_pig)),
             as.numeric(logLik(fit_zip))),
  AIC    = c(AIC(fit_pois), AIC(fit_pig), AIC(fit_zip))
)
comparison
```

Both mixed-Poisson models dominate the Poisson, and the zero-inflated PIG is
preferred, reflecting the excess zeros.

## Fitted versus empirical PMF

```{r pmf, fig.width = 7, fig.height = 5}
xv <- 0:30
emp <- tabulate(y + 1, nbins = max(xv) + 1)[seq_along(xv)] / length(y)

pmf_pig <- dpig(xv, mu = p_pig$mu, sigma = p_pig$sigma)
pmf_zip <- dzipig(xv, mu = p_zip$mu, sigma = p_zip$sigma, pi = p_zip$pi)

plot(xv, emp, type = "h", lwd = 3, col = "grey70",
     main = "Empirical vs fitted PMF", xlab = "Visits", ylab = "Probability")
lines(xv + 0.15, pmf_pig, type = "h", lwd = 2, col = "steelblue")
lines(xv + 0.30, pmf_zip, type = "h", lwd = 2, col = "firebrick")
legend("topright", c("Empirical", "PIG", "ZIPIG"), lwd = 2,
       col = c("grey70", "steelblue", "firebrick"))
```

## Randomized quantile residuals

For a discrete model a randomized quantile residual draws
$u \sim \mathrm{Unif}(F(y-1), F(y))$ and returns $\Phi^{-1}(u)$; if the model
is correct these are standard normal.

```{r rqr, fig.width = 7, fig.height = 4}
rqr <- function(y, pfun) {
  Fup <- pfun(y); Flo <- pfun(y - 1); Flo[y == 0] <- 0
  qnorm(runif(length(y), Flo, Fup))
}

set.seed(1)
r_pig <- rqr(y, function(q) ppig(q, mu = p_pig$mu, sigma = p_pig$sigma))
r_zip <- rqr(y, function(q) pzipig(q, mu = p_zip$mu, sigma = p_zip$sigma,
                                   pi = p_zip$pi))

par(mfrow = c(1, 2))
qqnorm(r_pig, main = "PIG", pch = 20, col = "grey50"); qqline(r_pig, col = "red")
qqnorm(r_zip, main = "ZIPIG", pch = 20, col = "grey50"); qqline(r_zip, col = "red")
par(mfrow = c(1, 1))
```

## Other fitting interfaces

The same distribution is available through the univariate maximum-likelihood
fitter `fitegpd()`, and as `gamlss` / `bamlss` families:

```{r other}
fit_uni <- fitegpd(y, family = "pig")
fit_uni$estimate

## PIG()/ZIPIG() (gamlss.dist) and pig_bamlss()/zipig_bamlss() provide the
## same distribution for distributional-regression frameworks.
```

The `dpig`, `ppig`, `qpig` and `rpig` functions (and their `zipig` counterparts)
give the density, distribution, quantile and simulation functions directly.
See also `vignette("gpig-counts")` for the heavier-tailed generalised PIG.
